---
title: "Linear Regression"
author: "Miguel Conde"
date: "3 de mayo de 2017"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```


# Key concepts

# Basics
Let's begin with just one predictor (X) for an outcome (Y).

If X and Y are highly correlated, we may assume that the true relationship between X and Y obeys a model like this:

$$
Y_i = \beta_0 + \beta_1 X_i + \epsilon_i
$$

Our work is to find out estimators of the true values $\beta_0$ and $\beta_1$ in order to estimate $Y_i$ as:

$$
\hat{Y}_i = b_0 + b_1 X_i
$$

So the residual $e_i$ accounts for:

$$
e_i = Y_i - \hat{Y}_i
$$
Meaning that $e_i$ is an estimator of $\epsilon_i$:

$$
Y_i = \beta_0 + \beta_1 X_i + \epsilon_i = b_0 + b_1 X_i + e_i = \hat{Y}_i + e_i
$$

# Model Estimation
**Ordinary least squares (*OLS*) regression**: The estimated model coefficients $b_0$ and $b_1$ are those which *minimise the sum of the squared residuals*:

$$
\sum_i{(Y_i - \hat{Y}_i)^2} = \sum_i{e_i^2}
$$

For an univariate linear regression:

$$
b_1 = Cor(Y, X) \frac{S_X}{S_Y} = \frac{Cov(Y, X)}{S_Y}
$$
and:
$$
b_0 = \bar{Y} - b_1 \bar{X}
$$
$S_X$ and $S_Y$ are empirical standard deviations:

$$
S_X^2 = \frac{1}{n-1}\sum_{i=1}^n(X_i - \bar{X})^2
$$
$\bar{X}$ is the empirical mean:

$$
\bar{X} = \frac{1}{n} \sum_{i=1}^nX_i
$$
The empirical covariance is:

$$
Cov(Y, X) = \frac{1}{n-1}\sum_{i=1}^n(Y_i - \bar{Y})(X_i - \bar{X})
$$

And the empirical correlation:
$$
Cor(Y, X) = \frac{Cov(Y, X)}{S_YS_X}
$$

## Example
```{r}
library(UsingR)
data(diamond)

library(highcharter)
hchart(diamond, "point", hcaes(x = carat, y = price))

```
These will be our predictor X and outcome Y:
```{r}
X <- diamond$carat
Y <- diamond$price
```

Fitting the model in R is easy:
```{r eval = FALSE}
m_fit <- lm (formula = price ~ carat, data = diamond)
```

But using our terminology:
```{r}
m_fit <- lm (Y ~ X)
summary(m_fit)
```

Let's calculate the coefficients step by step:
```{r}
n <- length(X)

m_X <- mean(X)
m_X
sum(X) / n

m_Y <- mean(Y)
```


```{r}
S_X <- sd(X)
S_X^2
var(X)
sum((X - m_X)^2) / (n - 1)

S_Y <- sd(Y)
```

```{r}
Cov_Y_X <- cov(X, Y)
Cov_Y_X
sum((X - m_X)*(Y - m_Y)) / (n - 1)
```
```{r}
Cor_Y_X <- cor(X, Y)
Cor_Y_X
Cov_Y_X / S_X / S_Y
```

And, finally:
```{r}
b_1 <- Cor_Y_X * S_Y / S_X
b_1
```

```{r}
b_0 <- m_Y - b_1 * m_X
b_0
```

Let's check our results:
```{r}
coef(m_fit)
```

```{r}
head(residuals(m_fit))
```

```{r}
head(Y - fitted(m_fit))
```

[`broom`](https://cran.r-project.org/web/packages/broom/index.html) is an interesting library:

*Convert statistical analysis objects from R into tidy data frames, so that they can more easily be combined, reshaped and otherwise processed with tools like 'dplyr', 'tidyr' and 'ggplot2'. The package provides three S3 generics: **tidy**, which summarizes a model's statistical findings such as coefficients of a regression; **augment**, which adds columns to the original data such as predictions, residuals and cluster assignments; and **glance**, which provides a one-row summary of model-level statistics.*


```{r}
library(broom)
head(augment(m_fit))
```

And this is our model visualization:

```{r}
library(dplyr)
fit <- arrange(augment(m_fit), X)
head(fit)
```

```{r}
hc <- highchart() %>% 
  hc_add_series(name = "Diamond", type = "point", data = diamond, 
                hcaes(x = carat, y = price)) %>%
  hc_add_series(name = "Fitted", type = "line", data = fit,
                hcaes(x = X, y = .fitted), id = "fit") %>%
  hc_add_series(type = "arearange", data = fit, 
                hcaes(x = X, low = .fitted - 2*.se.fit,
                      high = .fitted + 2*.se.fit),
                linkedTo = "fit") %>%
  hc_xAxis(title = list(text = "Carat")) %>%
  hc_yAxis(title = list(text = "Price")) %>%
  hc_title(text = "Price / Carat")
hc
```


# Multiple linear regression
Let's try now with multiple predictors, $X_1, X_2,...,X_K$. Our regression model will be:

$$
Y_i = \beta_0 + \sum_{j=1}^k{\beta_i X_{ij}} + \epsilon_i
$$



And we must estimate:
$$
\hat{Y}_i = b_0 + \sum_{i=1}^k{b_i X_{ij}}
$$
Of course, the residual $e_i$ is:

$$
e_i = Y_i - \hat{Y}_i
$$

Using matrices we can write:

$$
Y =  X \beta + \epsilon \\
\hat{Y} = XB
$$


We'll use these data:
```{r}
data("GaltonFamilies")
str(GaltonFamilies)
```

## In R
```{r}
m_fit2 <- lm(childHeight ~ father + mother, data = GaltonFamilies)
summary(m_fit2)
```

## Analytical fit

$$
B = (X^TX)^{-1}X^TY
$$

```{r}
N <- nrow(GaltonFamilies)

Y <- GaltonFamilies$childHeight
X1 <- GaltonFamilies$father
X2 <- GaltonFamilies$mother

mY <- matrix(Y, ncol = 1, nrow = length(Y))
mX <- matrix(c(rep(1, N), X1, X2), ncol = 3, nrow = N)

## OLS
mB <- solve( t(mX) %*% mX ) %*% t(mX) %*% mY
mB

coef(m_fit2)
```



# Model performance

## Whole model performance

### The $R^2$ value
We'd like that the sum of the squared residuals
$$
SS_{res} = \sum_i{(Y_i - \hat{Y}_i)^2}
$$

should be small, specifically in comparison to the total variability of the outcome:
$$
SS_{tot} = \sum_i{(Y_i - \bar{Y}_i)^2}
$$
With our regression:
```{r}
Y_hat <- mX %*% mB

SS_resid <- sum((Y - Y_hat)^2)
SS_resid
```

```{r}
SS_tot <- sum((Y - mean(Y))^2)
SS_tot
```

What we’d like to do is to convert these two fairly meaningless numbers
into one number. 

If we define:
$$
R^2 = 1- \frac{SS_{res}}{SS_{tot}}
$$

* $R^2 = 1$  if the regression model makes no errors in predicting
the data, as $SS_{res}=0$.

* $R^2 = 0$ if the model can't account at all for any of the outcome variability, i.e., $SS_{res}=SS_{tot}$.

```{r}
R_squared <- 1 - SS_resid/SS_tot
R_squared
```

So, the $R^2$ value interpretation is the following: it is the proportion of the variance in the outcome variable that can be accounted for by the predictors. In this case, the predictors explains `r round(100*R_squared, 2)`% of the variance in the outcome.

#### Regression and correlation
For a simple regression with just 1 predictor, the square of the Pearson correlation coefficient, $r^2$, is identical to the $R^2$ value.

Going back to our univariate example:
```{r}
X <- diamond$carat
Y <- diamond$price

cor(X, Y)^2
```
```{r}
s_m_fit <- summary(m_fit)
s_m_fit$r.squared
```
The same number.

#### The adjusted $R^2$ value
Adding more predictors into the model will always make $^2$ to increase. To avoid this behavioour, we use *adjusted* $R^2$. If we have $K$ predictors and $N$ observations:

$$
adj.R^2 = 1 - (\frac{SS_{res}}{SS_{tot}} \times \frac{N-1}{N-K-1})
$$
Note that $N-1$ are the residuals degrees of freedom and $N - K - 1$ are the total degrees of freedom, as we'll soon see.

However, the adjusted $R^2$ can't be interpreted as the proportion of vaiance in the outcome variable that is explained by the model.

### Hypothesis tests I: the F-statistic
We're going to try this hypothesis: there is no relationship between the predictors and the outcome:

$$
H_0:Y_i = b_0 + e_i
$$
The alternative hypothesis, of course, is:
$$
\hat{Y}_i = b_0 + \sum_{i=1}^k{b_i X_{ij}} + e_i
$$
To test these hypothesis, the trick is divide up the total variance $SS_{tot}$ into the sum of the residuals variance $SS_{resid}$ and the regression model variance$SS_{mod}$:

$$
SS_{tot} = SS_{mod} + SS{res}
$$
Hence:
$$
SS_{mod} = SS_{tot} - SS_{res} = \sum_{i=1}^N(Y_i - \bar{Y}) - \sum_{i=1}^N(Y_i - \hat{Y}_i) = \sum_{i=1}^N( \hat{Y}_i - \bar{Y})
$$
To convert the sums of squares into mean squares we divide by the degrees of freedom:

$$
MS_{mod} = \frac{SS_{mod}}{df_{mod}} \\
MS_{res} = \frac{SS_{res}}{df_{res}}
$$
Where:

* $df_{mod} = K$, the number of predictors
* $df_{tot} = N - 1$, $N$ observations minus 1 outcome mean $\bar{Y}$
* $df_{res} = df_{tot} - df_{mod} = N - K - 1$

Now we calculate the F-statistic:

$$
F = \frac{MS_{mod}}{MS_{res}}
$$
Large F values indicate that the null hypothesis is performing poorly in comparison to the alternative hypothesis. And the associated p-value allow us to reject or not $H_0$.



```{r}
s_m_fit$fstatistic
```



## Hypothesis tests II: tests for individual coefficients.
Passing the F-statistic test (i.e., rejecting the null) doesn't imply that the model is good, while failing this test is a pretty strong indicator that the model has problems.

# Confidence intervals

# Assumptions of linear regression and how to check them

# Regression model selection

