---
title: 't-tests: comparing two means'
author: "Miguel Conde"
date: "5 de mayo de 2017"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# The one-sample $z$-test

Let's suppose we are considering some population that can be described by a normal distribution whose parameters are perfectly knwon to us: mean $\mu_0 = 67.5$ and standard deviation $\sigma_0 = 10$.


```{r}
mu_0    <- 67.5
sigma_0 <- 10
```

Consider now *another* population, also described by a normal distribution and whose standard deviation is perfectly known to us, $\sigma = 9.5$

We now take some sample from this second population, let's say we obtain:
```{r}
sample_data <- c(50, 60, 60, 64, 66, 
                 66 ,67, 69, 70, 74, 
                 76, 76, 77, 79, 79,
                 79, 81, 82, 82, 89)
```

And calculate some summary statistics:
```{r}
mean(sample_data)
sd(sample_data)
```


We wonder if the mean of the second population is or not equal to the first's or it's been just a matter of chance (due to the small sample size) that $\bar{X} > \mu$.

This is rarelly a real situation because, in the real world, we probably won't be aware of the true distribution of the population (i.e., \sigma will be unknown).

Hence, our **null hypothesis** is:
$$
H_0: \mu = \mu_0 = 67.5
$$

And the **alternative hypothesis**:
$$
H_a: \mu \neq \mu_0 
$$


How can we discriminate between $H_0$ and $H_a$?

First, let's measure the distance between $\mu_0$ and $\bar{X}$:

$$
\bar{X} - \mu_0
$$

Is this quantity is pretty close to 0, the null hypothesis is quite likely to be true. But what does "pretty close to 0" mean?

If the null hypothesis is true then:

1 - We are certain that the second population is normal with mean $\mu_0$ and standar deviation $\sigma_1$.

$$
X \sim Normal(\mu_0, \sigma^2 )
$$

2 - By means of the CLT, the distribution of the sample means $\bar{X}$ is also normal with standard deviation (aka *standard error*):
$$
SE(\bar{X}) = \frac{\sigma}{\sqrt{N}}
$$

$$
\bar{X} \sim N(\mu_0, SE(\bar{X}))
$$

Well, let's measure now the same distance but in "number of standar errors" units.This is called a **standard score** (in this case, a $z$-score):

$$
z_{\bar{X}} = \frac{\bar{X} - \mu_0}{SE(\bar{X})} = \frac{\bar{X} - \mu_0}{\frac{\sigma}{\sqrt{N}}}
$$
Carefully notice that this $z$-score has a *standard normal distribution*:

$$
z_{\bar{X}} \sim Normal(0, 1)
$$

In other words, regardless of what scale the original data are on, the $z$-statistic itself always has the same interpretation: *it’s equal to the number of standard errors that separate the observed sample mean $\bar{X}$ from the population mean $\mu_0$ predicted by the null hypothesis*.

And the $\alpha$% critical regions for $z$-test are always the same, as illustrated in this table:

| desired $\alpha$ level | 2-sided test | 1-sided                           |
|:----------------------:|:------------:|:---------------------------------:|
| .1                     | `r -qnorm(.1/2)`  | `r qnorm(0.1, lower.tail = FALSE)`    |
| .05                    | `r -qnorm(.05/2)`  | `r qnorm(.05, lower.tail = FALSE)`   |
| .01                    | `r -qnorm(.01/2)`  | `r qnorm(.01, lower.tail = FALSE)`   |
| .001                   | `r -qnorm(.001/2)`  | `r qnorm(.001, lower.tail = FALSE)` |

## Example
### Step by step
```{r}
mu_0  <- 67.5
sigma <- 9.5

N <- length(sample_data)

sem <- sigma / sqrt(N) # Standard error = standard deviation of the mean

sample_mean <- mean(sample_data)

z_score <- (sample_mean - mu_0) / sem
z_score
```
Let's calculate the p-value.

```{r}
upper_area <- pnorm(q = z_score, lower.tail = FALSE)
upper_area

lower_area <- pnorm(q = -z_score)
lower_area
```

```{r}
p_value <- upper_area + lower_area
p_value
```

### With R
Study carefully `z_test()`
```{r}
source("z_Test.R")
z_test(x = sample_data , mu = mu_0, stdev = sigma)
```

We could sum up this way:

*"With a mean of `r sample_mean` in the sample , and assuming a true population standard deviation of `r sigma`, we can conclude that the sample population mean have significantly different statistics scores to the reference population ($z = `r round(z_score, 3)`$, $N = `r N`$, $p = `r round(p_value, 3)`$)."*